---
title: "Getting started with satbathy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with satbathy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

## Overview

**satbathy** estimates water depth in shallow, low-turbidity coastal waters
from multispectral satellite imagery using the empirical log-ratio transform of
Stumpf, Holderied and Sinclair (2003). The idea is simple: two visible bands
(typically blue and green) are attenuated at different rates as depth
increases, so the ratio of their log-transformed reflectance is, over a useful
range, linearly related to depth. That linear relationship is calibrated with a
handful of in situ soundings and then applied to the whole image.

This vignette walks through the full workflow on the bundled example dataset, a
Landsat scene with in situ bathymetry from Mahahual, on the Mexican Caribbean
coast.

```{r setup, message = FALSE}
library(satbathy)
library(terra)

img  <- rast(system.file("extdata", "example_sr.tif",     package = "satbathy"))
pts  <- vect(system.file("extdata", "example_points.gpkg", package = "satbathy"))
deep <- vect(system.file("extdata", "example_deepmask.gpkg", package = "satbathy"))

img
```

The image holds five surface-reflectance bands named `coastal`, `blue`,
`green`, `red` and `nir`. The points carry a `depth` attribute (metres, negative
below the surface).

## 1. Selecting clear-water dates

The ratio transform performs best in clear water, so the first practical step is
to pick the least turbid dates from an image time series. `turbidity_index()`
implements the index of Frohn and Autrey (2009), `(green + red) / blue`; lower
values mean clearer water. Here we simply summarise it for the single example
scene.

```{r turbidity}
turb <- turbidity_index(img)
global(turb, "mean", na.rm = TRUE)
```

In practice you would compute this index for every date and choose the minima
(for example with a moving-average smoother over the time series).

## 2. Masking land and clouds

Only water pixels should enter the model. When a sensor quality band is not
available, a threshold on the Normalized Difference Water Index works well:

```{r mask}
wm <- water_mask(img, green = "green", nir = "nir", threshold = 0)
img_water <- mask(img, wm, maskvalues = c(FALSE, NA))
```

**Sensor-specific preprocessing.** The bundled image is already surface
reflectance, cropped and land/cloud masked. Producing such an image from raw
archives depends on the sensor and is intentionally left outside the package,
but the recipe is:

* **Landsat 8/9 (Collection 2 L2).** Scale the `SR_B*` bands with the
  `REFLECTANCE_MULT_BAND` / `REFLECTANCE_ADD_BAND` values in the MTL metadata,
  and build masks from the `QA_PIXEL` bit flags (water bit; dilated cloud,
  cirrus, cloud and cloud-shadow bits).
* **Sentinel-2 (L2A).** Use the Scene Classification Layer (`SCL`) to keep water
  and discard clouds and shadows; reflectance is the `B*` bands divided by the
  quantification value.
* **PlanetScope / SuperDove (L2 SR).** There is no reliable water class in the
  UDM2 mask, so derive a water mask from the NDWI as above with `ndwi()` /
  `water_mask()`.

## 3. Optional glint and smoothing

Sun glint can be removed with the image-based method of Hedley et al. (2005),
which regresses each visible band on the near-infrared over optically deep
water (supplied as a polygon mask). A low-pass filter then reduces residual
speckle. Both steps are optional and, in very calm, uniform waters, make little
difference.

```{r glint}
img_gc <- correct_glint(img_water, vis_bands = c("blue", "green"),
                        nir_band = "nir", deep_mask = deep)
img_gc <- lowpass_filter(img_gc)
```

## 4. The log-ratio transform

```{r ratio}
ratio <- log_ratio(img_gc, blue = "blue", green = "green", n = 10000)
plot(ratio, main = "Log-ratio (blue/green)")
```

Darker offshore tones correspond to higher ratio values, i.e. greater depth.

## 5. Calibration

Pair the ratio with the in situ depths, then decide how deep the signal is
reliable. `optimal_max_depth()` refits the model for a decreasing sequence of
maximum-depth thresholds: goodness of fit rises as the linear range fills in and
then drops once saturated (too-deep) points are included.

```{r maxdepth}
tab <- extract_ratio(ratio, pts, depth_field = "depth")

md <- optimal_max_depth(tab)
plot(md$max_depth, md$r2, type = "b",
     xlab = "Maximum depth (m)", ylab = "Adjusted R-squared")

best <- md$max_depth[which.max(md$r2)]
best
```

The scatter makes the saturation explicit: beyond the penetration limit the
ratio no longer increases with depth.

```{r saturation}
plot(tab$ratio, tab$depth, pch = 20, col = "grey50",
     xlab = "Ratio", ylab = "Depth (m)")
abline(h = best, lty = 2)
```

Now fit the model over its valid range and inspect it:

```{r fit}
fit <- fit_bathymetry(tab, max_depth = best)
fit
plot(fit)
```

## 6. Depth map and accuracy

Apply the calibrated model to the ratio raster to obtain a depth map:

```{r predict}
depth_map <- predict_bathymetry(fit, ratio)
plot(depth_map, main = "Satellite-derived bathymetry (m)")
```

For an honest error estimate, hold out independent points within the valid depth
range and score them with `evaluate_bathymetry()`:

```{r evaluate}
valid <- tab[tab$depth > best, ]
set.seed(1)
i <- sample(nrow(valid), round(0.7 * nrow(valid)))
train <- valid[i, ]
test  <- valid[-i, ]

fit_tr <- fit_bathymetry(train)
evaluate_bathymetry(fit_tr, test)
```

## References

- Frohn RC, Autrey BC (2009). *Water quality assessment in the Ohio River using
  new indices for turbidity and chlorophyll-a with Landsat-7 imagery.* U.S. EPA.
- Hedley JD, Harborne AR, Mumby PJ (2005). *Simple and robust removal of sun
  glint for mapping shallow-water benthos.* International Journal of Remote
  Sensing 26(10):2107-2112.
- McFeeters SK (1996). *The use of the Normalized Difference Water Index (NDWI)
  in the delineation of open water features.* International Journal of Remote
  Sensing 17(7):1425-1432.
- Stumpf RP, Holderied K, Sinclair M (2003). *Determination of water depth with
  high-resolution satellite imagery over variable bottom types.* Limnology and
  Oceanography 48(1, part 2):547-556.
